% Abstract. Rewritten to (i) lead with the physical question, (ii) report
% HELD-OUT evidence rather than the training-set R^2 that headed the original
% draft, and (iii) attribute the gain to the predictive latent *family* rather
% than to the objective in isolation, pending the matched-architecture controls.

We ask whether a predictive latent training objective yields a reduced-order
state for parametric vortex-gust airfoil interactions that is more useful for
forward-time forecasting than a reconstructive one. The data are numerical
simulations of a NACA~0012 airfoil at angle of attack $\alpha = 14^\circ$ and
chord Reynolds number $\vRe = 5000$, perturbed by a discrete Taylor vortex
parametrised by gust strength $G$, core diameter $D$, and wall-normal offset
$Y$, spanning $84$ cases partitioned into $226$ training, $28$ validation,
$42$ in-distribution test, and $24$ out-of-distribution extrapolation
encounters. Three encoder families are compared at matched latent dimension:
a joint-embedding predictive architecture (JEPA), an observable-augmented
reconstructive autoencoder in the lineage of Fukami \& Taira, and a proper
orthogonal decomposition (POD) basis. A single autoregressive transformer
predictor, training recipe, parametric conditioning on $(G,D,Y)$, and probe
family are held fixed across all three, so that the encoder is the only thing
that varies. We judge each reduced state not by snapshot reconstruction but by
\emph{forward physical closure}: whether the predictor rollout keeps six
physical observables, namely the lift and drag coefficients, the wall-normal
impulse, the wake enstrophy, and the positive and negative circulation, close
to the simulation out to a horizon of $H = 16$ frames after gust impact. On
held-out encounters the predictive latent leads on the wake observables in point
estimate across both modes: its representational wake-enstrophy error is $2.4$ to
$3.0$ times lower than the reconstructive and linear baselines, its
representational wake-enstrophy $R^2$ is positive ($0.75$) where theirs are
negative, and it is the only latent whose forward forecast of wake enstrophy
stays above the predict-the-mean floor. The gains on the force observables are
smaller but consistent, and the linear basis remains competitive on the
integrated impulse. Bootstrap intervals over the $42$ held-out encounters are
wide for the per-observable closure, so the decisive and statistically robust
evidence is mechanistic. The mechanism is a
latent-drift diagnostic: under autoregressive rollout the reconstructive latent
leaves its training manifold by an order of magnitude at $H = 16$, whereas the
predictive and linear latents remain within it, so that the wake observables
that are hidden from a reconstruction objective become recoverable from a
predictive one. A conditioning-only floor shows that the held-out advantage is
not explained by the parametric conditioning the predictor shares across
baselines. A matched-architecture, matched-supervision $2 \times 2$ control
isolates the cause: the predictive objective improves held-out wake closure over
the reconstructive objective at both a CNN and a CNN$+$ViT encoder, so the gain
is the objective and not the architecture, while a no-wake-head ablation shows
the auxiliary wake supervision is a necessary component of the configuration that
achieves it. The advantage is thus a property of the predictive objective trained
with wake supervision, not of the objective or the architecture alone. The predictor is the latent dynamics function
required by model-based control, and we discuss the pathway to closed-loop
deployment and its current limits.
