% Appendix: architecture detail, anti-collapse regularisation, and uncertainty
% quantification. Describes only what is used on the present (v2) dataset: the
% SIGReg regulariser with its participation-ratio safeguard, and the three
% held-out uncertainty signals. The SIGReg/VICReg/PLDM comparison was a smaller-
% scale development study and is deliberately NOT claimed here.

\paragraph{Architecture detail.} The encoder is a hybrid convolutional stem
followed by a small vision transformer. Three downsampling convolutional stages
map the single-channel $192 \times 96$ vorticity field to a $24 \times 12$ feature
map at $256$ channels, that is $288$ spatial tokens, which a six-layer
pre-norm transformer of width $256$ with eight heads processes; the latent is read
from a prepended class token through a one-layer projection whose output passes a
BatchNorm, the layer the characteristic-function regulariser below requires. The
encoder is unconditional and carries no information about the gust parameters. The
predictor is a six-layer autoregressive transformer of width $384$ with sixteen
heads and dropout $0.1$, with a causal mask and rotary temporal positions. The
gust parameters $\cvec = (G, D, Y)$ and the within-encounter phase enter only the
predictor, through adaptive layer normalisation initialised to the identity
(AdaLN-Zero), so that at initialisation the predictor is the unconditioned
dynamics and the conditioning is learned as a modulation. Training minimises the
one-step latent prediction loss plus a scheduled-sampling rollout term and the
regulariser below, with AdamW and bf16 mixed precision.

\paragraph{Anti-collapse regularisation.} Because the encoder and predictor are
trained end to end on latent-space prediction with no decoder, the latent is held
away from a collapsed solution by an explicit regulariser rather than by a
reconstruction term. We use SIGReg, the characteristic-function (Epps--Pulley)
regulariser of the from-pixels joint-embedding line \cite{lewm,lejepa}, with $M$
random one-dimensional projections of the latent and a small weight ($0.01$) in
the loss. As a safeguard the participation ratio of the latent is monitored every
thousand iterations, with an automatic fall-back to a variance-covariance
objective \cite{vicreg} if it indicates collapse while the parameter probe is
still uninformative; on the runs reported here the participation ratio stays
healthy and the fall-back does not trigger.

\paragraph{Uncertainty quantification.} Every held-out coefficient of
determination and mean absolute error in the paper carries three independent
uncertainty signals, following the reporting protocol fixed with the split. The
first is a bootstrap confidence interval from $2000$ resamples over the held-out
test encounters, which is the dominant source of scatter at the present sample
size ($n = 42$ for test\_b, $n = 24$ for test\_c) and is wide for the individual
wake observables, as noted in \S\ref{sec:res_closure}. The second is the variance
across three encoder retrains from independent random seeds, reported as the
standard deviation in the objective-times-architecture controls of
\S\ref{sec:res_controls}. The third is a five-fold cross-validation over cases on
the read-out (probe) step, which guards against a probe that overfits the
particular held-out split. Together these approximate what a full five-fold
case-level cross-validation of the entire pipeline would buy, at a fraction of the
training cost; a single split with bootstrap intervals is the community standard
for this class of study.

\paragraph{Training-set fit.} For reference, Table~\ref{tab:b1_closure_train_r2}
reports the in-distribution training fit that the held-out closure of
\S\ref{sec:res_closure} must be read against. The predictive latent attains the
highest training fit (mean $R^2 = 0.835$ at $d = 64$), but the ordering of the
families is established by the held-out closure of Table~\ref{tab:closure}, not
by this in-distribution number.

\begin{table}[t]
  \centering
  \caption{Training-set fit, for reference only. Coefficient of determination
  $R^2$ for predicting the six observables from the rolled-out latent ($H = 16$)
  evaluated \emph{on the training distribution}. This quantifies in-distribution
  fit and is \emph{not} the headline; held-out closure is in
  Table~\ref{tab:closure}. Best per column in bold; row mean at right.}
  \label{tab:b1_closure_train_r2}
  \small
  \begin{tabular}{l c c c c c c c c}
    \toprule
    Baseline & $d$ & $C_L$ & $C_D$ & $I_y$ & wake-enst.\ & circ.\ pos.\ & circ.\ neg.\ & mean \\
    \midrule
    JEPA   & 64 & \textbf{0.864} & \textbf{0.802} & 0.562 & \textbf{0.934} & \textbf{0.922} & \textbf{0.927} & \textbf{0.835} \\
    JEPA   & 32 & 0.863 & 0.795 & 0.490 & 0.898 & 0.901 & 0.897 & 0.808 \\
    \midrule
    Fukami & 3  & 0.553 & 0.408 & 0.191 & 0.079 & 0.130 & 0.033 & 0.232 \\
    Fukami & 32 & 0.695 & 0.596 & 0.278 & 0.333 & 0.369 & 0.313 & 0.430 \\
    Fukami & 64 & 0.671 & 0.589 & 0.335 & 0.277 & 0.336 & 0.352 & 0.427 \\
    \midrule
    POD    & 16 & 0.763 & 0.720 & 0.514 & 0.574 & 0.582 & 0.527 & 0.613 \\
    POD    & 32 & 0.763 & 0.748 & 0.690 & 0.589 & 0.618 & 0.621 & 0.671 \\
    POD    & 64 & 0.641 & 0.656 & \textbf{0.799} & 0.373 & 0.388 & 0.506 & 0.561 \\
    \bottomrule
  \end{tabular}
\end{table}

\paragraph{Paired closure across all observables.} Table~\ref{tab:paired_closure}
reports the per-encounter paired comparison of \S\ref{sec:res_closure} for all six
observables in both modes, the source of the wake-enstrophy significance cited in
the main text. Pairing cancels the encounter-to-encounter difficulty that widens
the marginal intervals of Table~\ref{tab:closure}.

\begin{table}[t]
  \centering
  \caption{Paired held-out closure, predictive ($d = 64$) against reconstructive
  ($d = 64$) on the $n = 42$ test\_b encounters at $H = 16$, for the
  representational mode (probe on the simulation-encoded latent) and the forecast
  mode (Markov rollout from impact). $\Delta\mathrm{err}$ is the per-encounter
  mean of (reconstructive $-$ predictive) absolute error in each observable's own
  units, so a positive value favours the predictive latent; the $95\%$ interval
  is a $2000$-resample paired bootstrap, and the sign test counts the encounters
  on which the predictive error is the smaller of the pair. Wake enstrophy
  separates the families in both modes by both the bootstrap interval and the sign
  test; all six observables are shown, not a selected subset.}
  \label{tab:paired_closure}
  \small
  \begin{tabular}{l l r c l}
    \toprule
    Mode & Observable & $\Delta\mathrm{err}$ & $95\%$ CI & sign test \\
    \midrule
    repr.\ & $C_L$          & $+0.364$ & $[+0.070, +0.643]$ & $25/42$, $p = 0.14$ \\
    repr.\ & $C_D$          & $+0.039$ & $[-0.052, +0.131]$ & $20/42$, $p = 0.68$ \\
    repr.\ & $I_y$          & $+0.52$  & $[-0.028, +1.08]$  & $25/42$, $p = 0.14$ \\
    repr.\ & wake-enst.\    & $+43.1$  & $[+23.5, +66.0]$   & $31/42$, $p = 1.4{\times}10^{-3}$ \\
    repr.\ & circ.\ pos.\   & $+0.256$ & $[-0.117, +0.673]$ & $20/42$, $p = 0.68$ \\
    repr.\ & circ.\ neg.\   & $+1.19$  & $[+0.567, +1.93]$  & $30/42$, $p = 4.0{\times}10^{-3}$ \\
    \midrule
    forecast & $C_L$        & $-0.021$ & $[-0.206, +0.152]$ & $20/42$, $p = 0.68$ \\
    forecast & $C_D$        & $+0.066$ & $[+0.011, +0.128]$ & $26/42$, $p = 0.082$ \\
    forecast & $I_y$        & $+0.264$ & $[-0.164, +0.666]$ & $26/42$, $p = 0.082$ \\
    forecast & wake-enst.\  & $+32.0$  & $[+10.8, +54.8]$   & $27/42$, $p = 4.4{\times}10^{-2}$ \\
    forecast & circ.\ pos.\ & $-0.018$ & $[-0.319, +0.297]$ & $19/42$, $p = 0.78$ \\
    forecast & circ.\ neg.\ & $+0.555$ & $[+0.067, +1.11]$  & $18/42$, $p = 0.86$ \\
    \bottomrule
  \end{tabular}
\end{table}
