% Introduction. Related work (former Section 2) is folded in here, in the JFM
% idiom (no standalone "Related work" section). The SIGReg-vs-PLDM "second paper"
% that headed Section 1.2 of the original draft has been removed; the regulariser
% choice is now a methods detail (Sec. 3) and an appendix. Contributions are
% reframed around forward physical closure, with the objective-isolation claim
% explicitly deferred to the controls.

Small uncrewed and micro air vehicles increasingly operate in airspace where
the gust velocity is comparable to or larger than their flight speed, such as
urban canyons, mountainous terrain, and the wakes of buildings and ships
\cite{jones2022,mohamed2023gusts}. In this regime the gust ratio $G$, the ratio of
the characteristic gust velocity to the free-stream velocity, exceeds unity,
and the interaction between a discrete vortex gust and a lifting surface
produces large, fast transients in the aerodynamic loads that conventional
gust models do not capture \cite{fukami2024control,fukami2025prf}. The response
is governed by the timing of the vortex impact relative to the airfoil's own
shedding, by the wall-normal offset of the vortex core, and by the gust
strength and core size. Even at the moderate chord Reynolds number
$\vRe = 5000$, the post-stall flow at $\alpha = 14^\circ$ is massively
separated, and the gust-induced disturbance interacts nonlinearly with the
natural shedding cycle \cite{fukami2025prf}. Resolving the full parametric
envelope by simulation is expensive, so reduced-order models (ROMs) of the
encounter-to-encounter and case-to-case variability are an active line of work,
and a reduced model is also the object a controller would plan against.

Recent ROMs for separated and extreme-aerodynamic flows share a common
ingredient: a reconstruction-trained autoencoder whose bottleneck is shaped by
an auxiliary aerodynamic observable. Fukami \& Taira \cite{fukami2023} trained
a convolutional autoencoder jointly with a lift-prediction head and found that
the auxiliary loss reduced the intrinsic dimension of the discovered manifold
for extreme vortex-gust flows from roughly ten to three. The same
observable-augmentation idea was extended across multi-source turbulent data
\cite{fukami2025}, and was used by Solera-Rico et al. \cite{solera2024} with a
$\beta$-variational autoencoder and a transformer ROM on a vortex-gust airfoil
dataset of the same family considered here. Across this lineage the
reconstruction term anchors the latent geometry to the instantaneous flow
field, and an observable-prediction term orients it along aerodynamically
relevant directions. A recurring theme of the geometric-analysis literature,
however, is that the latent coordinates of an autoencoder are unique only up to
a diffeomorphism, so reconstruction leaves the geometry of the latent largely
unconstrained, which complicates any downstream task that relies on that
geometry, including dynamics, interpolation, and comparison
\cite{tran2026,kelshaw2024}.

A complementary route to a reduced state is to drop the reconstruction term
altogether and to train the encoder and a predictor end to end on
latent-space forecasting, with an explicit anti-collapse regulariser in place
of the decoder. This is the joint-embedding predictive architecture (JEPA)
\cite{lecun2022}: an encoder maps the input to a representation, a predictor
advances that representation in time, and the loss never reconstructs the
field. Two consequences follow. The encoder is free to discard detail that is
not predictive of the future, and the predictor cannot exploit pixel-level
shortcuts because no field-space signal reaches it. The image and video
instantiations \cite{ijepa,vjepa2} and the recent from-pixels variants that
remove the moving-target encoder in favour of a characteristic-function
regulariser \cite{lewm,lejepa} or a variance-covariance objective
\cite{pldm,vicreg} have so far been evaluated on gridworld and toy visual
domains. Their behaviour on a low-intrinsic-dimension physics dataset, where
the data sit on a roughly five- to ten-dimensional manifold, is uncharacterised.

Reconstruction-trained nonlinear encoders compress aggressively but yield latent
dynamics that are harder to forecast than an energy-optimal linear basis, a
compactness-versus-forecast tradeoff documented for controlled wake flows in a
companion study under review \cite{solerarico_compactness_underreview}. We show
this tradeoff is a consequence of the reconstruction objective: a predictive
latent recovers nonlinear compactness without surrendering forecast stability.

This places the present model within the action-conditioned world-model
framework articulated by \cite{lecun2022}: a system is observed through an
encoder into an abstract state, an intervention is applied, and a predictor
forecasts the outcome of that intervention in the representation space rather
than in the space of the raw observation. The defining property of such a model
is that it does not reconstruct every detail of the outcome; it predicts only in
the abstract space. The vortex gust is the intervention. A discrete vortex of
strength $G$, size $D$, and offset $Y$ is released into the post-stall wake, and
the encounter is the outcome of that intervention; the encoder maps the
vorticity field to an unconditional latent state, and the predictor advances
that state conditioned on the gust parameters. We use the gust as an observed,
exogenous intervention rather than an agent's chosen action, so the model is the
interventional world model a controller would use, not yet a controller; and
because we condition on the true generative parameters but do not perform
counterfactual identification, we keep to the interventional reading and do not
claim a full causal model. The reconstructive and linear baselines we compare
against are, in this framing, the objects the world-model view argues against: a
reconstruction-trained autoencoder is pushed toward reproducing every detail of
the field, and a proper orthogonal decomposition is an energy-optimal
reconstruction basis, neither of which is constrained to make the latent state
forward-predictable. The drift mechanism of \S\ref{sec:res_drift} is the
empirical consequence.

The question this paper addresses is therefore not which encoder reconstructs
the flow best, but which produces a reduced state that remains useful when
propagated forward in time. We treat this as a fluid-mechanics question with a
physical answer: a reduced state for gust-airfoil forecasting should be judged
by whether its forward evolution keeps physically meaningful observables close
to the simulation, and by whether the latent trajectory it generates stays on
the manifold the model was trained on. We make three contributions.

First, we set up a controlled comparison in which a predictive encoder (JEPA),
a reconstructive encoder in the Fukami lineage, and a POD basis are evaluated
at matched latent dimension under a single shared autoregressive predictor,
training recipe, parametric conditioning on $\cvec = (G,D,Y)$, and probe
family, so that forecast quality is attributable to the encoder alone
(\S\ref{sec:methods}). The figure of merit is forward physical closure of the
rollout to six observables at horizon $H = 16$ (\S\ref{sec:res_closure}). On
held-out encounters the predictive latent closes the spatially distributed wake
observables markedly better than the reconstructive and linear baselines, while
the linear basis remains competitive on the integrated impulse.

Second, we identify the mechanism. A latent-drift diagnostic shows that the
reconstructive rollout leaves its training manifold by an order of magnitude at
$H = 16$, whereas the predictive and linear rollouts remain within it
(\S\ref{sec:res_drift}). We strengthen this from a single distance into a
geometric and topological statement: the predictive latent organises the
encounter as a closed cycle whose topology survives rollout, and its
latent metric tracks the physical transport geometry of the flow more
faithfully than the reconstructive latent does.

Third, we are explicit about what the comparison does and does not isolate. The
predictive encoder differs from the reconstructive baseline not only in its
objective but also in its architecture and in its auxiliary observable
supervision. We therefore phrase the result as a property of the
\emph{predictive latent family}, report a conditioning-only floor that rules
out the shared parametric conditioning as the explanation (\S\ref{sec:res_closure}),
and set out the matched-architecture and matched-auxiliary controls required to
attribute the gain to the predictive objective in isolation
(\S\ref{sec:res_controls}). We close by relating the predictor to the latent
dynamics function of model-based control, including a deployment pilot that does
not yet meet its tolerances and is reported as a limitation rather than a result
(\S\ref{sec:discussion}).
