% Discussion. Former Section 8, restructured and de-diaried. The closed-loop
% pilot is reported as a limitation/future pathway, not a contribution. The
% objective-isolation caveat is stated plainly. The latent non-uniqueness framing
% (Tran et al. 2026) is used to explain *why* reconstruction leaves the geometry
% free.

\subsection{What the predictive latent retains, and why}
\label{sec:disc_mechanism}

The central finding is about dynamic closure, not reconstruction. Under a shared
predictor, parametric conditioning, and probe family, the predictive latent
family keeps the forward evolution of six physical observables closer to the
simulation than the reconstructive and linear families do, and the advantage is
concentrated on the spatially distributed wake observables that integrate over
the shedding region. The mechanism is geometric. A reconstruction objective
fixes the latent only up to a diffeomorphism \cite{tran2026,kelshaw2024}, so the
decoder is free to place codes anywhere it finds convenient and nothing
constrains the latent geometry to be predictable in time; under autoregressive
rollout the reconstructive latent therefore leaves its own training manifold,
and a probe queried on that drifted state fails for reasons unrelated to the
predictor's one-step accuracy. The predictive objective, regularised against
collapse, instead organises the latent so that the encounter is a closed cycle
that survives rollout and so that latent distance tracks the physical transport
geometry of the flow. Iterating the predictor therefore does not leave the
manifold, the property that data-driven state-space and Koopman models of
coherent-state dynamics on invariant manifolds rely on
\cite{constanteamores2024}. The manifold departure of the reconstructive rollout
is the
parametric-gust counterpart of the catastrophic-divergence tail reported for
reconstruction-based latent dynamics in controlled wakes
\cite{solerarico_compactness_underreview}; the predictive objective is the common
cure. The wake observables are the sharpest discriminator precisely because
they can change while the scalar forces do not: a representation can carry the
lift signature of an encounter without carrying the vorticity redistribution
that produces it, and the reconstructive latent does exactly that, recovering
the impact lift poorly from its own oracle while the predictive latent recovers
it well.

This organisation survives a halving of the latent dimension, which is the basis
for reading it as a low-dimensional predictive state rather than an artefact of
capacity. At $d = 32$ the predictive latent matches the $d = 64$ latent on
representation quality, the held-out wake-enstrophy representation $R^2$ is $0.74$
against $0.75$, and on every mechanism diagnostic: it traces the same single
persistent cycle rather than fragmenting (\S\ref{sec:res_drift}), its latent
metric aligns with the physical transport geometry to within the $d = 64$ margin
(rank correlation $0.61$ against $0.63$), it returns to the baseline limit cycle
under rollout (\S\ref{sec:res_param_phase}), it retains the same large-scale wake
structure (\S\ref{sec:res_physical}), and its rollout-drift ratio is $0.86$
against $0.85$ (Table~\ref{tab:latent_drift}). The one quantity that degrades is
the in-distribution wake forecast, where the $H = 16$ closure $R^2$ falls from
$0.45$ to $0.21$ (Table~\ref{tab:b1_r2_heldout}): halving the dimension costs
forecast sharpness on the wake observables while leaving the geometric mechanism
intact. On the out-of-distribution test\_c set the $d = 32$ latent is in fact
marginally the more robust of the two, $0.45$ against $0.33$ on wake enstrophy.
The physics the predictive objective organises into the latent, the closed
encounter cycle and its transport-consistent metric, is therefore a
low-dimensional object already present at $d = 32$, and the additional dimensions
at $d = 64$ buy in-distribution forecast precision rather than a qualitatively
different representation. The parameter content of the two latents is consistent
with this reading: the dominant gust-strength axis is encoded as well at $d = 32$
as at $d = 64$, while the core diameter and the wall-normal offset, the axes the
data resolve least, are the ones whose recovery degrades when the dimension is
halved, the same axes whose forward closure the extra dimensions sharpen.

\subsection{What the comparison does and does not establish}
\label{sec:disc_caveat}

We are deliberate about the scope of the claim. The predictive encoder differs
from the reconstructive baseline in three respects at once: the training
objective, the CNN$+$ViT architecture, and the wake supervision. The
conditioning-only floor removes the shared parametric conditioning as an
explanation, and the matched-architecture, matched-auxiliary $2 \times 2$
controls of \S\ref{sec:res_controls} separate the remaining two. They give a
two-part answer that we state in full. The predictive objective does improve
forward wake closure over the reconstructive objective at both a CNN and a
CNN$+$ViT encoder with the auxiliary heads matched, and the two architecture
columns do not separate, so the gain is the objective and not the architecture.
The wake supervision, however, is a necessary part of the configuration that
achieves it: a predictive model with the wake head removed collapses below the
predict-the-mean floor on wake closure. The honest claim is therefore that the
predictive objective improves forward closure \emph{when trained with wake
supervision}, not that the objective in isolation does, and not that the wake
head attached to an arbitrary objective does, since the reconstructive cells
carry the same head and do not reach the predictive closure. The linear basis,
for its part, remains genuinely useful: it is the most competitive family on the
integrated impulse and is amplitude-accurate in reconstruction, so the result is
not that POD is obsolete but that it does not produce a latent whose rollout
closes the wake observables.

\subsection{Limitations}
\label{sec:disc_limits}

Three limitations bound the claims. First, the wall-normal offset $Y$ is weakly
resolved because the training mass concentrates near $Y \approx 0$; this is a
data limitation, and $Y$-conditional statements should be read with it in mind.
Second, the $|G| = 4$ extrapolation degrades for a physical reason, not merely a
statistical one: the encoder observes a single mid-plane vorticity slice, and at
$|G| = 4$ the true dynamics become three-dimensional \cite{fukami2025prf}, so
test\_c is best read as the observability boundary of a mid-plane representation
rather than as a parametric-generalisation claim. Third, the visualisation
decoder is a diagnostic on the frozen encoder, never part of the predictive
loss; a reader expecting sharp reconstructions from a predictive latent will be
disappointed, which is the explicit trade the objective makes. To these we add,
as a fourth, the co-necessity of the wake supervision established in
\S\ref{sec:disc_caveat}: the predictive objective improves forward closure in the
configuration tested, with the wake head as a necessary component, not as a
property of the objective stripped of all auxiliary supervision.

\subsection{Pathway to model-based control, and a pilot that does not yet close}
\label{sec:disc_control}

The predictor is, by construction, the latent dynamics function a model-based
controller plans against, and three properties of the architecture support that
use: closure holds out to the planning horizon relevant to the impact-instant
load, the latent is two to three orders of magnitude cheaper to evolve than the
full field, and the rollout stays on the training manifold, which is the
property that determines whether long-horizon plans remain meaningful since a
planner queries the predictor on states reached by its own rollout. Because the
encoder is unconditional, the latent itself is the observable a controller has
access to, and the gust parameters enter only the predictor; replacing the
parametric conditioning with an actuation channel is a change of input rather
than of architecture, and the phase-amplitude reading of \S\ref{sec:res_param_phase}
connects the predictor directly to the sensitivity-function control designed for
these flows \cite{fukami2024control}. In the action-conditioned world-model
reading of \S\ref{sec:introduction}, the present intervention is the gust; an actuation
channel enters at the same point, the predictor conditioning, and the same
forward-closure machinery applies with the control as the action rather than the
disturbance. The predictor is the world model a controller would use; it is not
itself a controller, and the pilot below shows the gap.

We also report, as a limitation rather than a result, a closed-loop pilot that
chains a wall-pressure state estimator, the predictor rollout, and an
observable probe. A temporal-convolutional estimator recovers the latent from a
pre-impact pressure window well (held-out $R^2 \approx 0.87$ at eight sensors),
and at matched latent dimension the predictive state is markedly more recoverable
from the wall than the reconstructive or linear latents ($R^2 = 0.89$ versus
$0.58$ and $0.32$ at $d = 64$; Appendix~\ref{app:sensing}), the deployment mirror
of the representation result.
But the assembled loop does not meet the tolerances set in advance: at the
design horizon only a small fraction of held-out encounters fall within the
$C_L$ and $I_y$ tolerance bands, and none of the pre-registered gates are met.
The bottleneck is the rollout rather than the estimator, since the within-
tolerance fraction is essentially unchanged when the oracle latent is
substituted for the estimated one. The honest reading is that the representation
and the pressure observability are in place but the rollout needs reinforcement,
through more frequent re-estimation from pressure during the rollout or a longer
scheduled-sampling horizon, before a usable controller follows. The estimator,
sensor-placement, and closed-loop details are deferred to Appendix~\ref{app:sensing}.

\subsection{Outlook}
\label{sec:disc_outlook}

The contribution is narrow by design: a controlled comparison on a single
parametric flow, anchored to forward physical closure rather than to
in-distribution reconstruction, with a drift mechanism that explains the gap and
a set of controls that bound its interpretation. The immediate next step is to
run those controls and the horizon, topology, transport, and phase-amplitude
analyses set out above, which together would either promote the claim to the
predictive objective in isolation or sharpen it to the predictive family. Beyond
that, applying the same matched-predictor closure protocol to other parametric
flow datasets would test how general the predictive-versus-reconstructive trade
reported here actually is.
